---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Before analyzing the data, we need to load the data into R. The assignment
instructions inform us that the input data contains three columns: `steps`,
`date`, and `interval`. Furthermore the instructions indicate that the dataset
is stored in a comma-separated-value (CSV) file and that there are a total of
17,568 observations in it. The input data is in the file `activity.zip`. From
the extension, we will assume that it is in compressed format.

Just to make sure that our assumption is correct and to get a sense of the
input data, we will take a quick peek at a few records from the input file.
For this task we could use external programs like `head` or `tail`. However,
while these are usually available on *nix systems, they are not available on
standard Windows systems (unless added on later by Cygwin or msys etc.).

To remain platform-agnostic, instead of using `read.csv()` from the base R
system we will use the `fread()` function from the `data.table` package. Also,
the data loading code assumes that open source 7-Zip archive management
software is available in the in the search path of the OS.

```{r}
# Set the working directory and get the contents of the .zip file
setwd("~/Coursera/Data Science Specialization/5. Reproducible Research/RepData_PeerAssessment1")
cat(paste(system("7za l activity.zip", intern = TRUE), "\n"))

# Load the data.table package
library(data.table)

# Read the data into a data.table with default fread() options
dt = fread("7za e -so activity.zip activity.csv 2>nul")

# Take a look at the data
str(dt)
dt
```

The instructions indicate that the variable `interval` is the identifier for
5-minute intervals. The first intervals start with the series [0, 5, 10, ...],
and if it were a continuous series during the day, we would expect the
following:

```{r}
# Number of intervals in a day
intervals_per_hour = 60/5
intervals_per_day = 24 * intervals_per_hour

# Sequence of intervals for one day
data.table(seq(from = 0, by = 5, length.out = intervals_per_day))
```

The tail end of the source dataset doesn't match this vector. In fact, it looks
more like the time of the day in 24-hour format, but without the ":" separator.
To confirm this, we can take a quick look at the end of the hour and the end of
the day:

```{r}
# Check out the end of the first hour
dt[(intervals_per_hour - 1):(intervals_per_hour + 2)]

# Check out the 24-hour boundary between the first and second day
dt[(intervals_per_day - 1):(intervals_per_day + 2)]
```

It looks like the `interval` column is actually the time of the day, but it is
formatted as a number: e.g. `55` is followed by `100`; `2355` is followed by
`0`, etc. In case we need a properly formatted `interval` and a proper date and
time variable in later steps, we will transform them and keep them in a new
Date/Time column.

```{r, message=FALSE}
# Load lubridate package for easy handling of dates/times
library(lubridate)

# Reformat date and time and convert to date/time class
dt[, interval := sub("^(..)", "\\1:", sprintf("%04d", interval))]
dt[, timestamp := ymd_hm(paste(date, interval))]
dt[, date := ymd(date)]
str(dt)
```


## What is mean total number of steps taken per day?

To summarize data by group, we can use the `aggregate()` function from the base
R system, or utilize the modern interfaces provided in `data.table` or `dplyr`
packages. For this assignment, we will use `data.table` interface.

```{r}
# Get the steps by date, ignore missing values
steps_by_date = dt[, .( total_steps = sum(steps, na.rm = TRUE) ), date]

# Load the ggplot2 library
library(ggplot2)

# Plot with ggplot
ggplot(steps_by_date, aes(x = date, y = total_steps)) +
  geom_bar(stat = "identity") +
  xlab("Date") +
  ylab("Steps") +
  ggtitle("Total Number of Steps Taken Each Day")

# mean steps taken
mean_steps = steps_by_date[, mean(total_steps, na.rm = TRUE)]

# median steps taken
median_steps = steps_by_date[, median(total_steps, na.rm = TRUE)]
```

During the study period, after removing the missing data, the mean steps taken
per day was **`r prettyNum(mean_steps, big.mark = ",")`** and the median steps
taken per day was **`r prettyNum(median_steps, big.mark = ",")`**.



## What is the average daily activity pattern?

To show the daily activity pattern, we will plot the average steps taken at
each interval over the duration of the study.

```{r, message=FALSE}
# Average of steps taken by interval over the study period
steps_by_interval = dt[, .(average_steps = sum(steps, na.rm = TRUE)), interval]

# Info on the interval with the max average steps
max_step_index     = steps_by_interval[, which.max(average_steps)]
max_steps          = steps_by_interval[max_step_index, average_steps]
max_steps_interval = steps_by_interval[max_step_index, interval]

# There are too many intervals to be plotted: pick the top of the hour to label
# the tick marks
labels = steps_by_interval[seq(1, length(interval), 12), interval]

# For comma formatted y axis labels
require(scales)

# Line graph of average activity during the day
activity = ggplot(steps_by_interval, aes(x = interval, y = average_steps, group = 1)) +
  geom_line() +
  ggtitle("Average Daily Activity Pattern") +
  xlab("Intervals") +
  ylab("Average Steps Taken") +
  scale_x_discrete(breaks=labels, labels=as.character(labels)) +
  scale_y_continuous(labels = comma) +
  theme(axis.text.x=element_text(angle=90))

# Now, add the info about the max steps to the plot and show it
activity +
  geom_segment(
    linetype = "longdash",
    colour = "blue",
    aes(x = max_step_index, y = -Inf, xend = max_step_index, yend = max_steps)
  ) +
  geom_text(
    x = max_step_index, y = max_steps,
    label = prettyNum(max_steps, big.mark = ","),
    hjust=0, vjust=0
  ) +
  geom_text(
    x = max_step_index, y = 0,
    label = max_steps_interval,
    hjust=0.25, vjust=1.25,
    angle = 90
  )
```

On the average, the maximum number of steps during a 5 minute interval were
**`r prettyNum(max_steps, big.mark = ",")`**, taken at
**`r max_steps_interval`**.



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
